---
title: "Dataset_preprocessing"
author: "Dillon Corvino"
date: "05/12/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Human 
# data downloaded from https://zenodo.org/records/7562864

```

### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)

# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- FALSE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())

# load packages
library("Seurat")
library("SeuratExtend")

# Load the future package
library(future)

# Set the maximum size of globals that can be exported to 50 GB
options(future.globals.maxSize = 100 * 1024^3)  # 50 GB in bytes
getOption("future.globals.maxSize")


# Ensure ggplot2 is the correct version - requires restarting Rsession and loading packages again
ggplot2_switch_version(desired_version = "3.5.0")


# genes of interest 
adenosine.genes <- c("NT5E", "ENTPD1", "ADORA1", "ADORA2A", "ADORA2B", "ADORA3")

# load processed data
if(quick.load){
  
    haniffa.seurat <- LoadSeuratRds("saves/haniffa_seurat.rds")

}

```


## Pre-processing {.tabset}

### Load dataset
```{r load_dataset}

if(long.compute){

    # Create output dir
    output.dir <- paste0("results/Overview/")
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
    }
      
      
      
  input.path <- "data/Zenodo_7562864_dataset/SC-RNA-DATA"
  
  list.dirs(input.path)
  
  # Haniffa data 
  haniffa.path <- "/HANIFFA-DATA/RDS-Files/"
  haniffa.healthy <- readRDS(paste0(input.path, haniffa.path, "healthy_skin.sc_data_HNF.RDS"))
  haniffa.psoriasis <- readRDS(paste0(input.path, haniffa.path, "HNF_SC_RNA_PSORIASIS_DATA.RDS"))
  #haniffa.data <- LoadH5Seurat(paste0(input.path, haniffa.path, "submission.h5seurat"))
  
head(haniffa.healthy@meta.data)
head(haniffa.psoriasis@meta.data)

# are healthy samples also present in psoriasis dataset or unique 
healthy.result <- gsub(".*-", "", rownames(haniffa.healthy@meta.data))
healthy.result <- unique(healthy.result)

psoriasis.result <- gsub(".*-", "", rownames(haniffa.psoriasis@meta.data))
psoriasis.result <- unique(psoriasis.result)
healthy.result %in% psoriasis.result
psoriasis.result %in% healthy.result




meta.vars <- c("Status", "Site", "Tissue", "Enrichment", "Location", "donor_id", "Sex", "Age", "stage", "final_clustering")

for(i in seq_along(meta.vars)){

p1 <- DimPlot_scCustom(haniffa.healthy,
                       group.by = meta.vars[i],
                       label = T,
                       raster = T,
                       pt.size = 1) + NoLegend() + ggtitle("Healthy_dataset ", meta.vars[i])

p2 <- DimPlot_scCustom(haniffa.psoriasis,
                       group.by = meta.vars[i],
                       label = T,
                       raster = T,
                       pt.size = 1) + NoLegend() + ggtitle("Psoriasis_dataset ", meta.vars[i])

print(p1|p2)
ggsave(paste0(output.dir, "Haniffa_dataset_", meta.vars[i], ".pdf"))

}


# Sullivan data 
 # sullivan.path <- "/SULLIVAN-DATA/RDS-Files/"
 # sullivan.data <- readRDS(paste0(input.path, sullivan.path, "LEAN_SAMPLES_SCRNA.rds"))

#head(sullivan.data@meta.data)
#DimPlot2(sullivan.data, group.by = "CellType")



# Travis data 
travis.path <- "/TRAVIS-DATA/RDS-Files/"
#travis.data <- readRDS(paste0(input.path, travis.path, "Travis_Data_Seurat_object.RDS"))
travis.NM.skin <- readRDS(paste0(input.path, travis.path, "travis_NM_skin_scRNA.RDS"))
travis.psoriasis <- readRDS(paste0(input.path, travis.path, "TRAVIS_PSORIASIS_ONLY_DATA.RDS"))


#head(travis.data@meta.data)
head(travis.NM.skin@meta.data)
head(travis.psoriasis@meta.data)


meta.vars <- c("CellType", "Array", "Condition", "Sample", "SampleCondition", "Specific_CellType", "SingleR_Labels")

for(i in seq_along(meta.vars)){

p1 <- DimPlot_scCustom(travis.NM.skin,
                       group.by = meta.vars[i],
                       label = T,
                       raster = T,
                       pt.size = 1) + NoLegend() + ggtitle("Healthy_dataset ", meta.vars[i])

p2 <- DimPlot_scCustom(travis.psoriasis,
                       group.by = meta.vars[i],
                       label = T,
                       raster = T,
                       pt.size = 1) + NoLegend() + ggtitle("Psoriasis_dataset ", meta.vars[i])

print(p1|p2)

ggsave(paste0(output.dir, "Travis_dataset_", meta.vars[i], ".pdf"))

}

    
    # merge haniffa datasets
    haniffa.seurat <- scCustomize::Merge_Seurat_List(list_seurat = list("Healthy" = haniffa.healthy, 
                                                                      "Psoriasis" = haniffa.psoriasis), 
                                                  add.cell.ids = c("Healthy", "Psoriasis"))
    
    haniffa.seurat # 33,538 genes across 343,706 cells
    table(haniffa.seurat@meta.data$Status) # 195,739 healthy and 147,967 Psoriasis (some lesion some nonlesion)
    
    
    
    
# remove un-needed data objects
    rm(list=c("haniffa.healthy", "haniffa.psoriasis", "travis.NM.skin", "travis.psoriasis"))
    
    
    }

```



## QC and Filter {.tabset}

### QC & filter
```{r QC_and_filter}

if(long.compute){
  
  
    # Create output dir
    output.dir <- paste0("results/QC/")
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
      
    }
    
    
    # set idents to sampleID
    Idents(haniffa.seurat) <- haniffa.seurat$Status
    
    # set default assay 
    DefaultAssay(haniffa.seurat) <- "RNA"
    
    # Add mitochondrial and ribosomal gene percentages
    
    haniffa.seurat <- Add_Mito_Ribo(object = haniffa.seurat, species = "Human", overwrite = TRUE) 
    
    # Add cell complexity/novelty QC metric
   haniffa.seurat <- Add_Cell_Complexity(object = haniffa.seurat, overwrite = TRUE)
    
    # Setting cutoff values 
    feature.low <- 200
    feature.high <- 5000
    umi.low <- 1000
    umi.high <- 45000
    mito.high <- 10
    ribo.high <- 2
    complexity.high <- 0.75
    
    
    ########################
    # QC plots
    ########################
    if(do.plot){
      # nFeatures
      print(
        QC_Plots_Genes(seurat_object = haniffa.seurat, 
                       low_cutoff = feature.low, 
                       high_cutoff = feature.high,
                       pt.size = 0) 
      )
      
      dev.copy(pdf, paste0(output.dir, "genes.pdf"))
      dev.off()
      
      
      # UMI
      print(
        QC_Plots_UMIs(seurat_object = haniffa.seurat, 
                      low_cutoff = umi.low, 
                      high_cutoff = umi.high, 
                      y_axis_log = T,
                      pt.size = 0)
      )
      
      dev.copy(pdf, paste0(output.dir, "UMI.pdf"))
      dev.off()
      
      
      # Mito percentage
      print(
        QC_Plots_Mito(seurat_object = haniffa.seurat, 
                      high_cutoff = mito.high,
                      pt.size = 0)
      )
      
      dev.copy(pdf, paste0(output.dir, "mito.pdf"))
      dev.off()
      
      
      # Ribo percentage
      print(
        QC_Plots_Feature(seurat_object = haniffa.seurat,
                         feature = "percent_ribo",
                         high_cutoff = ribo.high,
                         pt.size = 0)
      )
      
      dev.copy(pdf, paste0(output.dir, "Ribo.pdf"))
      dev.off()
      
      
      # Complexity
      print(
        QC_Plots_Complexity(seurat_object = haniffa.seurat, 
                            high_cutoff = complexity.high,
                            pt.size = 0)
      ) 
      
      dev.copy(pdf, paste0(output.dir, "complexity.pdf"))
      dev.off()
      
    }
    
    # Filter dataset
    haniffa.seurat <- subset(haniffa.seurat, 
                                  subset = nFeature_RNA > feature.low & 
                                    nFeature_RNA < feature.high & 
                                    percent_mito < mito.high & 
                                    percent_ribo > ribo.high & 
                                    log10GenesPerUMI > complexity.high)
    
    if(do.plot){
      # nFeatures filtered matrix
      print(
        QC_Plots_Genes(seurat_object = haniffa.seurat, 
                       low_cutoff = feature.low, 
                       high_cutoff = feature.high,
                       pt.size = 0) #how many genes per cell? See where to set the cutoffs later
      )
      
      dev.copy(pdf, paste0(output.dir, "genes_filtered.pdf"))
      dev.off()
      
      
      # UMI filtered matrix
      print(
        QC_Plots_UMIs(seurat_object = haniffa.seurat, 
                      low_cutoff = umi.low, 
                      high_cutoff = umi.high, 
                      y_axis_log = T,
                      pt.size = 0)
      )
      
      dev.copy(pdf, paste0(output.dir, "UMI_filtered.pdf"))
      dev.off()
      
      # Ribo filtered percentage
      print(
        QC_Plots_Feature(seurat_object = haniffa.seurat,
                         feature = "percent_ribo",
                         high_cutoff = ribo.high,
                         pt.size = 0)
      )
      
      dev.copy(pdf, paste0(output.dir, "Ribo_filtered.pdf"))
      dev.off()
      
      
      
  
}

    
}
```

## Normalise and integrate {.tabset}

### Normalise & integrate GEX data
```{r Normalise_and_integrate_GEX_data}

if(long.compute){
  
  # set default assay 
  DefaultAssay(haniffa.seurat) <- "RNA"
  
  #####################################################
  # Normalise with scTransform and integrate data
  #####################################################
  # note error with more recent version of matrixStats resulting in "useNames = NA is defunct" error. downgrade to matrixStats version 1.1.0 as a workaround
  # https://github.com/satijalab/seurat/issues/7501#issuecomment-1854571904
  #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts  
  
  
  # note RNA assay is split into run level layers, below script will apply normalisation etc layer-wise
  haniffa.seurat[["RNA"]] <- split(haniffa.seurat[["RNA"]], f = haniffa.seurat$Status)

  # split by healthy and psoriasis datasets, integrate the two datasets together
  
  # Normalise and scale with SCTransform v2
  haniffa.seurat <- SCTransform(haniffa.seurat, 
                           method = "glmGamPoi", 
                           vars.to.regress = "percent_mito", 
                           vst.flavor = "v2",
                           verbose = TRUE) %>%
    RunPCA(reduction.name = "pca_sct") %>%
    RunUMAP(reduction = "pca_sct",
            dims = 1:30, 
            reduction.name = "umap.unintegrated.sct")
  
   
  DefaultAssay(haniffa.seurat) <- "SCT"
  
  # integrate layers using cca integration
  haniffa.seurat <- IntegrateLayers(object = haniffa.seurat,
                               orig.reduction = "pca_sct",
                               new.reduction = "integrated.cca.sct",
                               method = CCAIntegration,
                               normalization.method = "SCT",
                               verbose = TRUE)
  
  # find clusters and dim reduction using integrated data
  haniffa.seurat <- FindNeighbors(haniffa.seurat,
                             dims = 1:30,
                             reduction = "integrated.cca.sct") %>%
    FindClusters(resolution = 1.5, 
                 cluster.name = "clusters_cca_sct") %>%
    RunUMAP(reduction = "integrated.cca.sct",
            dims = 1:30, 
            reduction.name = "umap.cca.sct")
  

  if(do.plot){
    
    # create output dir
    output.dir <- paste0("results/integration/")
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
      
    }
    
    
    unintegrated.reduction <- "umap.unintegrated.sct"
    integrated.reduction <- "umap.cca.sct"
    title.var <- "SCT"
    
    
    # evaluate integration and dim reduction
    
    vars.plot <- c("seurat_clusters", "final_clustering", "Sex", "Tissue", "Status", "Enrichment", "donor_id", "Site")
    
    for(i in seq_along(vars.plot)){
      
      
      # remove "." from name
      name.var <- gsub("\\.", "_", vars.plot[i])
      
      before.int <- DimPlot(haniffa.seurat,
                            reduction = unintegrated.reduction, 
                            label = T,
                            group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("Unintegrated ", title.var, " Normalisation"))
      
      after.int <- DimPlot(haniffa.seurat,
                           reduction = integrated.reduction, 
                           label = T,
                           group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("CCA Integration using ", title.var, " Normalisation"))
      
      print(before.int | after.int)
      dev.copy(pdf, paste0(output.dir, "UMAP_before_after_integration", name.var, ".pdf"))
      dev.off()
      
      
      
    }
    
  }
  
}


```

### Save seurat object
```{r save_seurat_object}

if(long.compute){
  
  SaveSeuratRds(haniffa.seurat, file = "saves/haniffa_seurat.rds")
  
  # extract and save metadata 
  write.csv(haniffa.seurat@meta.data, "saves/metadata.csv")
}

```

### Overview
```{r Overview}

# create output dir
output.dir <- paste0("results/Dataset_overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

    vars.plot <- c("seurat_clusters", "final_clustering", "Sex", "Tissue", "Status", "Enrichment", "donor_id", "Site")

    
    for(i in seq_along(vars.plot)){
      
      p1 <- scCustomize::DimPlot_scCustom(haniffa.seurat,
                                          group.by = vars.plot[i], 
                                          label = TRUE,
                                          pt.size = 1,
                                          reduction = "umap.cca.sct") + 
        NoLegend() + 
        ggtitle(paste0(vars.plot[i]))
      
      print(p1)
      ggsave(filename = paste0(output.dir, "UMAP_", vars.plot[i], ".pdf"))
      
    }
    
    
    # Cluster distribution bar plot
    
    vars.plot <- c("seurat_clusters", "final_clustering", "Sex", "Tissue", "donor_id", "Site")
    
    for(i in seq_along(vars.plot)){
      
      p1 <- ClusterDistrBar(haniffa.seurat$Status, haniffa.seurat@meta.data[,vars.plot[i]], 
                            rev = T, 
                            percent = T, 
                            normalize = T, 
                            plot = T)
      
      print(p1)
      ggsave(filename = paste0(output.dir, "Clust_Distribution_", vars.plot[i], "_normalised.pdf"))
      
      
      p2 <- ClusterDistrBar(haniffa.seurat$Status, haniffa.seurat@meta.data[,vars.plot[i]], 
                            rev = T, 
                            percent = F, 
                            normalize = T, 
                            plot = T)
      
      print(p2)
      ggsave(filename = paste0(output.dir, "Clust_Distribution_", vars.plot[i], "_normalised_abs.pdf"))
      
      
      
    }
    
    
    
```



### Adenosine genes overview
```{r adenosine_genes}

# create output dir
output.dir <- paste0("results/Adenosine_genes_overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}





  ############
  # GEX data
  ############
  DefaultAssay(haniffa.seurat) <- "SCT"
  
  Idents(haniffa.seurat) <- "seurat_clusters"
table(Idents(haniffa.seurat))



# create pseudobulk for visualisation
pseudo.seurat <- AggregateExpression(haniffa.seurat, 
                                     assays = "SCT",
                                     return.seurat = TRUE,
                                     group.by = c("seurat_clusters"))
  
# set idents of pseudobulk object
Idents(pseudo.seurat) <- "seurat_clusters"

DoHeatmap(pseudo.seurat, 
          features = adenosine.genes, 
          draw.lines = FALSE,
          raster = FALSE) + NoLegend()
  
dev.copy(pdf, paste0(output.dir, "Heatmap_seurat_clusters_pseudobulk_adenosine_genes.pdf"))
dev.off()



ggplot2_switch_version(desired_version = "3.4.4")


# Plot Feature and Density plots iteratively over goi 
for(i in seq_along(adenosine.genes)){
  
  # feature plots
  print(
    scCustomize::FeaturePlot_scCustom(haniffa.seurat, 
                                      reduction = "umap.cca.sct", 
                                      features = adenosine.genes[i])
    
  )
  
  dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_", adenosine.genes[i], ".pdf"))
  dev.off()
  
  # density plots
  
  print(
    scCustomize::Plot_Density_Custom(haniffa.seurat, 
                                     reduction = "umap.cca.sct",
                                     features = adenosine.genes[i], 
                                     custom_palette = batlow.pal)
  )
  dev.copy(pdf, paste0(output.dir, "UMAP_DensityPlot_", adenosine.genes[i], ".pdf"))
  dev.off()
  
  # Vln plots
  
  print(
    scCustomize::VlnPlot_scCustom(haniffa.seurat, 
                                  features = adenosine.genes[i], 
                                  pt.size = 0, 
                                  group.by = "seurat_clusters") + 
      NoLegend()
  )
  
  dev.copy(pdf, paste0(output.dir, "VlnPlot_", adenosine.genes[i], ".pdf"))
  dev.off()
  
}


ggplot2_switch_version(desired_version = "3.5.0")


#x <- SeuratExtend::feature_percent(haniffa.seurat, "NT5E", DefaultAssay = "SCT")
#sum(x)


```





## Annotate clusters {.tabset}

### Differential gene expression GEX vs seurat_clusters
```{r DEG_analysis_seurat_clusters}

# create output dir
output.dir <- paste0("results/DEG_analysis_seurat_clusters/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

################################################
# Caclulate GEX DEGs for across seurat_clusters 
################################################

if(long.compute){
  
  # set assay
  DefaultAssay(seurat.human) <- "SCT"
  seurat.human <- PrepSCTFindMarkers(seurat.human)
  
  # DEG calculation using single-cells
  
  # set assay
  DefaultAssay(seurat.human) <- "SCT"
  
  # format cluster annotations
  seurat.human@meta.data$seurat_clusters <- factor(seurat.human@meta.data$seurat_clusters, levels = c(0:length(unique(seurat.human@meta.data$seurat_clusters))))
  
  # set idents
  Idents(seurat.human) <- "seurat_clusters"
  
  seurat_clusters.gex.de <- FindAllMarkers(seurat.human,
                                           assay = "SCT",
                                           only.pos = TRUE, 
                                           verbose = TRUE)
  
  # save to file 
  write.csv(seurat_clusters.gex.de, paste0(output.dir, "DEGs_wilcox_single_cell_SCT_seurat_clusters.csv"))
  
}else{
  
  seurat_clusters.gex.de <- read.csv(paste0(output.dir, "DEGs_wilcox_single_cell_SCT_seurat_clusters.csv"))
  
}




```


### DEG heatmaps - GEX and ADT - seurat_clusters
```{r DEG_heatmaps_seurat_clusters}

# create output dir
output.dir <- paste0("results/DEG_analysis_seurat_clusters/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(do.plot){
  
  
  ############
  # GEX data
  ############
  DefaultAssay(seurat.human) <- "SCT"
  
  # create pseudobulk for visualisation
  pseudo.seurat <- AggregateExpression(seurat.human, 
                                   assays = "SCT",
                                   return.seurat = TRUE,
                                   group.by = c("seurat_clusters"))
  
  # set idents of pseudobulk object
  Idents(pseudo.seurat) <- "seurat_clusters"
  
  # extract top 10 DEGs per cluster
  goi <- seurat_clusters.gex.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::group_by(cluster) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    dplyr::pull(gene)
  
  # heatmap of top 10 degs per cluster
  DoHeatmap(pseudo.seurat, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + NoLegend()
  
  dev.copy(pdf, paste0(output.dir, "Heatmaptop_10_DEGs_seurat_clusters_pseudobulk.pdf"))
  dev.off()
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(seurat_clusters.gex.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- seurat_clusters.gex.de %>%
      dplyr::filter(p_val_adj < 0.05) %>%
      dplyr::filter(cluster == cluster.id[i]) %>%
      dplyr::top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.seurat, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE)
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk__seurat_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
}

```








### Save seurat object
```{r save_seurat_object}

if(long.compute){
  
  SaveSeuratRds(HL.seurat, file = "saves/HL_seurat_Azimuth.rds")
  
  # extract and save metadata 
  write.csv(HL.seurat@meta.data, "saves/metadata.csv")
}else{
  
  HL.seurat <- LoadSeuratRds("saves/HL_seurat_Azimuth.rds")
}

```




