---
title: "visualisation"
author: "Dillon Corvino"
date: "05/12/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Human 
# data downloaded from https://zenodo.org/records/7562864

```

### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)

# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- FALSE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())

# load packages
library("Seurat")
library("SeuratExtend")

# Load the future package
library(future)

# Set the maximum size of globals that can be exported to 50 GB
options(future.globals.maxSize = 100 * 1024^3)  # 50 GB in bytes
getOption("future.globals.maxSize")


# Ensure ggplot2 is the correct version - requires restarting Rsession and loading packages again
ggplot2_switch_version(desired_version = "3.5.0")


# genes of interest 
adenosine.genes <- c("NT5E", "ENTPD1", "ADORA1", "ADORA2A", "ADORA2B", "ADORA3")

# load processed data
if(quick.load){
  
    haniffa.seurat <- LoadSeuratRds("saves/haniffa_seurat.rds")

    
    # create new broad cell annotation 
    # group together finer detail celltypes
    
    library(dplyr)

haniffa.seurat@meta.data <- haniffa.seurat@meta.data %>%
  mutate(Broad_celltypes = case_when(
    final_clustering %in% c("Proliferating_KC", "Differentiated_KC", "Undifferentiated_KC*", "Differentiated_KC*") ~ "Keratinocytes",
    final_clustering %in% c("LE1", "LE2") ~ "Langerhans_Epidermal",
    final_clustering %in% c("F1", "F2", "F3") ~ "Fibroblasts",
    final_clustering %in% c("VE1", "VE2", "VE3") ~ "Vascular_Endothelial",
    final_clustering %in% c("LC_1", "LC_2", "LC_3", "LC_4") ~ "Langerhans_Cells",
        final_clustering %in% c("Th") ~ "CD4T_Cells",
    final_clustering %in% c("Treg") ~ "Treg",
    final_clustering %in% c("Tc") ~ "CD8T_Cells",

    #final_clustering %in% c("Th", "Tc", "Treg") ~ "T_Cells",
        final_clustering %in% c("NK") ~ "NK_cells",
    final_clustering %in% c("ILC1_NK", "ILC1_3", "ILC2") ~ "NK_ILCs",

#    final_clustering %in% c("NK", "ILC1_NK", "ILC1_3", "ILC2") ~ "Innate_Lymphoid_Cells",
    final_clustering %in% c("Macro_1", "Macro_2", "Mono", "Inf_mono") ~ "Monocytes_Macrophages",
    final_clustering %in% c("DC1", "DC2", "MigDC") ~ "Dendritic_Cells",
    final_clustering %in% c("moDC_1", "moDC_2", "moDC_3") ~ "Monocyte_Derived_DC",
    final_clustering == "Plasma" ~ "Plasma_Cells",
    final_clustering %in% c("Pericyte_1_non_inflamm", "Pericyte_2_inflamm") ~ "Pericytes",
    final_clustering %in% c("Schwann1", "Schwann2") ~ "Schwann_Cells",
    final_clustering == "Mast_cell" ~ "Mast_Cells",
    final_clustering == "Melanocyte" ~ "Melanocytes",
    TRUE ~ "Other"
  ))

    
    
}

```


## General Overview {.tabset}

### Overview
```{r Overview}

# create output dir
output.dir <- paste0("results/Dataset_overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

    vars.plot <- c("seurat_clusters", "final_clustering", "Sex", "Tissue", "Status", "Enrichment", "donor_id", "Site")

    
    for(i in seq_along(vars.plot)){
      
      p1 <- scCustomize::DimPlot_scCustom(haniffa.seurat,
                                          group.by = vars.plot[i], 
                                          label = TRUE,
                                          pt.size = 1,
                                          reduction = "umap.cca.sct") + 
        NoLegend() + 
        ggtitle(paste0(vars.plot[i]))
      
      print(p1)
      ggsave(filename = paste0(output.dir, "UMAP_", vars.plot[i], ".pdf"))
      
    }
    
    
    # Cluster distribution bar plot
    
    vars.plot <- c("seurat_clusters", "final_clustering", "Sex", "Tissue", "donor_id", "Site")
    
    for(i in seq_along(vars.plot)){
      
      p1 <- ClusterDistrBar(haniffa.seurat$Status, haniffa.seurat@meta.data[,vars.plot[i]], 
                            rev = T, 
                            percent = T, 
                            normalize = T, 
                            plot = T)
      
      print(p1)
      ggsave(filename = paste0(output.dir, "Clust_Distribution_", vars.plot[i], "_normalised.pdf"))
      
      
      p2 <- ClusterDistrBar(haniffa.seurat$Status, haniffa.seurat@meta.data[,vars.plot[i]], 
                            rev = T, 
                            percent = F, 
                            normalize = T, 
                            plot = T)
      
      print(p2)
      ggsave(filename = paste0(output.dir, "Clust_Distribution_", vars.plot[i], "_normalised_abs.pdf"))
      
      
      
    }
    
    
    
```


###  Where is Adenosine pathway expressed
```{r Adenosine_overview}

# create output dir
output.dir <- paste0("results/adenosine_overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

library(ggplot2)
library(scCustomize)
library(gplots)
library(Seurat)


p <- scCustomize::DimPlot_scCustom(haniffa.seurat,
                                   group.by = "final_clustering", 
                                   label = TRUE,
                                   pt.size = 1,
                                   figure_plot = T,
                                   reduction = "umap.cca.sct") 
print(p)
ggsave(filename = paste0(output.dir, "UMAP_final_clustering.pdf"),
       plot = p, 
       width = 8,
       height = 6, 
       device = "pdf")

p <- scCustomize::DimPlot_scCustom(haniffa.seurat,
                                   group.by = "Broad_celltypes", 
                                   label = TRUE,
                                   figure_plot = T,
                                   pt.size = 1,
                                   reduction = "umap.cca.sct") 
print(p)
ggsave(filename = paste0(output.dir, "UMAP_Broad_celltypes.pdf"),
       plot = p, 
       width = 8,
       height = 6, 
       device = "pdf")


  
      ############
  # GEX data
  ############
  DefaultAssay(haniffa.seurat) <- "SCT"
  
  
Idents(haniffa.seurat) <- "Broad_celltypes"



# create pseudobulk for visualisation
pseudo.seurat <- AggregateExpression(haniffa.seurat, 
                                     assays = "SCT",
                                     return.seurat = TRUE,
                                     group.by = c("Broad_celltypes"))
  
# set idents of pseudobulk object
Idents(pseudo.seurat) <- "Broad_celltypes"

DoHeatmap(pseudo.seurat, 
          features = adenosine.genes, 
          draw.lines = FALSE,
          raster = FALSE) + NoLegend()
  
dev.copy(pdf, paste0(output.dir, "Heatmap_seurat_clusters_pseudobulk_adenosine_genes.pdf"))
dev.off()



ggplot2_switch_version(desired_version = "3.4.4")

Idents(haniffa.seurat) <- "Broad_celltypes"

# Plot Feature and Density plots iteratively over goi 
for(gene in adenosine.genes){
  
  # feature plots
  p <- scCustomize::FeaturePlot_scCustom(haniffa.seurat, 
                                         reduction = "umap.cca.sct", 
                                         features = gene)
  
  print(p)
  ggsave(filename = paste0(output.dir, "UMAP_featureplot_", gene, ".pdf"),
                           plot = p, 
                           width = 8,
                           height = 6, 
                           device = "pdf")
  
  
  # density plots
  
  p <- scCustomize::Plot_Density_Custom(haniffa.seurat, 
                                        reduction = "umap.cca.sct",
                                        features = gene, 
                                        custom_palette = batlow.pal)
  
  print(p)
  ggsave(filename = paste0(output.dir, "UMAP_DensityPlot_", gene, ".pdf"),
                           plot = p, 
                           width = 8,
                           height = 6, 
                           device = "pdf")
  # Vln plots
  
  p <- scCustomize::VlnPlot_scCustom(haniffa.seurat, 
                                     features = gene, 
                                     pt.size = 0, 
                                     group.by = "Broad_celltypes") + 
    NoLegend()
  
  print(p)
  
  ggsave(filename = paste0(output.dir, "VlnPlot_", gene, ".pdf"),
                           plot = p, 
                           width = 8,
                           height = 6, 
                           device = "pdf")
  
  
}


ggplot2_switch_version(desired_version = "3.5.0")



library(ggplot2)
library(dplyr)
library(SeuratExtend)

# Define a fixed color scheme for cell types
celltype_colors <- c(
  "CD4T_Cells" = "#E41A1C",  # Red
  "CD8T_Cells" = "#377EB8",  # Blue
  "Dendritic_Cells" = "#4DAF4A",  # Green
  "Fibroblasts" = "#984EA3",  # Purple
  "Keratinocytes" = "#FF7F00",  # Orange
  "Langerhans_Cells" = "#FFFF33",  # Yellow
  "Langerhans_Epidermal" = "#A65628",  # Brown
  "Mast_Cells" = "#F781BF",  # Pink
  "Melanocytes" = "#999999",  # Grey
  "Monocyte_Derived_DC" = "#66C2A5",  # Teal
  "Monocytes_Macrophages" = "#FC8D62",  # Salmon
  "NK_cells" = "#8DA0CB",  # Light Blue
  "NK_ILCs" = "#E78AC3",  # Light Pink
  "Other" = "#A6D854",  # Light Green
  "Pericytes" = "#FFD92F",  # Gold
  "Plasma_Cells" = "#E5C494",  # Beige
  "Schwann_Cells" = "#B3B3B3",  # Light Grey
  "Treg" = "#1B9E77",  # Dark Teal
  "Vascular_Endothelial" = "#D95F02"  # Dark Orange
)

# Loop over each gene in adenosine.genes
for(gene in adenosine.genes){
  
  # Compute percentage expression per cluster
  expr_data <- SeuratExtend::feature_percent(haniffa.seurat, gene, group.by = "Broad_celltypes", above = 0, DefaultAssay = "SCT")
  
  # Convert to a data frame
  expr_df <- as.data.frame(t(expr_data))
  colnames(expr_df) <- "Percent_Expressing"
  expr_df$Cell_Type <- rownames(expr_df)
  
  # Sort by decreasing expression
  expr_df <- expr_df %>% arrange(desc(Percent_Expressing))
  
  # Ensure colors are assigned correctly
  expr_df$Color <- celltype_colors[expr_df$Cell_Type]
  
  # convert to freq out of 100%
  expr_df$Percent_Expressing <- expr_df$Percent_Expressing*100
  
  # Create bar plot with consistent colors
  p <- ggplot(expr_df, aes(x = reorder(Cell_Type, Percent_Expressing), y = Percent_Expressing, fill = Cell_Type)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = celltype_colors) +  # Use consistent colors
    coord_flip() +
    labs(title = paste(gene, "Expression Across Cell Types"),
         x = "Cell Type",
         y = "Percentage of Cluster Expressing Gene") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")  # Hide legend to keep plots clean

  # Save plot to file
  ggsave(filename = paste0(output.dir, "Expression_", gene, ".pdf"), plot = p, width = 8, height = 6)
}

# loop over each gene but split and calculate freq expressing per disease status 
# Loop over each gene in adenosine.genes

library(ggplot2)
library(dplyr)
library(SeuratExtend)
library(patchwork)  # For side-by-side plots

# Create Status-CellType combined annotation
haniffa.seurat@meta.data$Status_celltype <- paste0(haniffa.seurat@meta.data$Status, "_", haniffa.seurat@meta.data$Broad_celltypes)

# Loop over each gene in adenosine.genes
for(gene in adenosine.genes){
  
  # Compute percentage expression per cluster
  expr_data <- SeuratExtend::feature_percent(haniffa.seurat, gene, group.by = "Status_celltype", above = 0, DefaultAssay = "SCT")
  
  # Convert to a data frame
  expr_df <- as.data.frame(t(expr_data))
  colnames(expr_df) <- "Percent_Expressing"
  expr_df$Status_celltype <- rownames(expr_df)
  
  # Extract Status and Cell Type
  expr_df <- expr_df %>%
    mutate(Status = ifelse(grepl("^Healthy_", Status_celltype), "Healthy", "Psoriasis"),
           Cell_Type = gsub("Healthy_|Psoriasis_", "", Status_celltype),
           Percent_Expressing = Percent_Expressing * 100)  # Convert to percentage
  
  # Sort by decreasing expression
  expr_df <- expr_df %>% arrange(desc(Percent_Expressing))
  
  # Separate data into Healthy & Psoriasis
  expr_healthy <- expr_df %>% filter(Status == "Healthy")
  expr_psoriasis <- expr_df %>% filter(Status == "Psoriasis")
  
  # Ensure consistent color mapping
  expr_healthy$Color <- celltype_colors[expr_healthy$Cell_Type]
  expr_psoriasis$Color <- celltype_colors[expr_psoriasis$Cell_Type]
  
  # Get max y-axis limit for consistency
  y_max <- max(expr_df$Percent_Expressing, na.rm = TRUE)

  # Create bar plots for Healthy
  p1 <- ggplot(expr_healthy, aes(x = reorder(Cell_Type, Percent_Expressing), y = Percent_Expressing, fill = Cell_Type)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = celltype_colors) +
    coord_flip() +
    labs(title = paste(gene, "- Healthy"), x = "Cell Type", y = "Percentage Expressing") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none") +
    ylim(0, y_max)

  # Create bar plots for Psoriasis
  p2 <- ggplot(expr_psoriasis, aes(x = reorder(Cell_Type, Percent_Expressing), y = Percent_Expressing, fill = Cell_Type)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = celltype_colors) +
    coord_flip() +
    labs(title = paste(gene, "- Psoriasis"), x = "", y = "") +  # Remove redundant labels
    theme_minimal(base_size = 14) +
    theme(legend.position = "none",
          axis.ticks.y = element_blank()) +
    ylim(0, y_max)

  # Combine plots side-by-side
  combined_plot <- p1 + p2 + plot_annotation(title = paste(gene, "Expression in Healthy vs Psoriasis"))

  # Save plot to file
  ggsave(filename = paste0(output.dir, "Expression_disease_seperated_", gene, ".pdf"), plot = combined_plot, width = 12, height = 6)
}


```

## TO DO
# make vln plots boxwiskers plots and add statistics, is sig diff or not - perform t test 

### healthy vs disease comparison
```{r healthy_vs_disease}

# create output dir
output.dir <- paste0("results/adenosine_overview_healthy_vs_disease/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

Idents(haniffa.seurat) <- "Broad_celltypes"
celltypes <- unique(Idents(haniffa.seurat))

for(celltype in celltypes){
  
  temp.seurat <- subset(haniffa.seurat, idents = celltype)
  
  p <- SeuratExtend::WaterfallPlot_v3(temp.seurat,
                                      features = adenosine.genes, 
                                      ident.1 = "Healthy", ident.2 = "Psoriasis", group.by = "Status",
                                      length = "logFC", 
                                      style = "bar") + ggtitle(paste0("Healthy vs Psoriasis - ", celltype))
  print(p)
  
  ggsave(filename = paste0(output.dir, "Waterfall_adenosine_path_", celltype, ".pdf"),
         plot = p, 
         width = 8,
         height = 6, 
         device = "pdf")
  
  # create violn plots of genes individually 
  for(gene in adenosine.genes){
    
    p <- VlnPlot_scCustom(temp.seurat, 
                          features = gene, 
                          group.by = "Status", 
                          pt.size = 0) + NoLegend() + 
      ggtitle(ggtitle(paste0(celltype, "-", gene)))
    print(p)
    
    ggsave(filename = paste0(output.dir, "Vlnplot_", gene, "_", celltype, ".pdf"),
           plot = p, 
           width = 8,
           height = 6, 
           device = "pdf")
    
  }
  
}




```


# TO DO 
# group adenosine pathway by production and receptor (producer or responder) 
# maybe make a unified adenosine pathway score 
# zoom into most interesting populations (either producers or responders)
# then query the difference in expression between healthy and disease and maybe abudnace of the population across healthy and disease 





### Adenosine genes overview
```{r adenosine_genes}

# create output dir
output.dir <- paste0("results/Adenosine_genes_overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}





  ############
  # GEX data
  ############
  DefaultAssay(haniffa.seurat) <- "SCT"
  
  Idents(haniffa.seurat) <- "seurat_clusters"
table(Idents(haniffa.seurat))



# create pseudobulk for visualisation
pseudo.seurat <- AggregateExpression(haniffa.seurat, 
                                     assays = "SCT",
                                     return.seurat = TRUE,
                                     group.by = c("seurat_clusters"))
  
# set idents of pseudobulk object
Idents(pseudo.seurat) <- "seurat_clusters"

DoHeatmap(pseudo.seurat, 
          features = adenosine.genes, 
          draw.lines = FALSE,
          raster = FALSE) + NoLegend()
  
dev.copy(pdf, paste0(output.dir, "Heatmap_seurat_clusters_pseudobulk_adenosine_genes.pdf"))
dev.off()



ggplot2_switch_version(desired_version = "3.4.4")


# Plot Feature and Density plots iteratively over goi 
for(i in seq_along(adenosine.genes)){
  
  # feature plots
  print(
    scCustomize::FeaturePlot_scCustom(haniffa.seurat, 
                                      reduction = "umap.cca.sct", 
                                      features = adenosine.genes[i])
    
  )
  
  dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_", adenosine.genes[i], ".pdf"))
  dev.off()
  
  # density plots
  
  print(
    scCustomize::Plot_Density_Custom(haniffa.seurat, 
                                     reduction = "umap.cca.sct",
                                     features = adenosine.genes[i], 
                                     custom_palette = batlow.pal)
  )
  dev.copy(pdf, paste0(output.dir, "UMAP_DensityPlot_", adenosine.genes[i], ".pdf"))
  dev.off()
  
  # Vln plots
  
  print(
    scCustomize::VlnPlot_scCustom(haniffa.seurat, 
                                  features = adenosine.genes[i], 
                                  pt.size = 0, 
                                  group.by = "seurat_clusters") + 
      NoLegend()
  )
  
  dev.copy(pdf, paste0(output.dir, "VlnPlot_", adenosine.genes[i], ".pdf"))
  dev.off()
  
}


ggplot2_switch_version(desired_version = "3.5.0")


#x <- SeuratExtend::feature_percent(haniffa.seurat, "NT5E", DefaultAssay = "SCT")
#sum(x)


```





## Annotate clusters {.tabset}

### Differential gene expression GEX vs seurat_clusters
```{r DEG_analysis_seurat_clusters}

# create output dir
output.dir <- paste0("results/DEG_analysis_seurat_clusters/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

################################################
# Caclulate GEX DEGs for across seurat_clusters 
################################################

if(long.compute){
  
  # set assay
  DefaultAssay(seurat.human) <- "SCT"
  seurat.human <- PrepSCTFindMarkers(seurat.human)
  
  # DEG calculation using single-cells
  
  # set assay
  DefaultAssay(seurat.human) <- "SCT"
  
  # format cluster annotations
  seurat.human@meta.data$seurat_clusters <- factor(seurat.human@meta.data$seurat_clusters, levels = c(0:length(unique(seurat.human@meta.data$seurat_clusters))))
  
  # set idents
  Idents(seurat.human) <- "seurat_clusters"
  
  seurat_clusters.gex.de <- FindAllMarkers(seurat.human,
                                           assay = "SCT",
                                           only.pos = TRUE, 
                                           verbose = TRUE)
  
  # save to file 
  write.csv(seurat_clusters.gex.de, paste0(output.dir, "DEGs_wilcox_single_cell_SCT_seurat_clusters.csv"))
  
}else{
  
  seurat_clusters.gex.de <- read.csv(paste0(output.dir, "DEGs_wilcox_single_cell_SCT_seurat_clusters.csv"))
  
}




```


### DEG heatmaps - GEX and ADT - seurat_clusters
```{r DEG_heatmaps_seurat_clusters}

# create output dir
output.dir <- paste0("results/DEG_analysis_seurat_clusters/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(do.plot){
  
  
  ############
  # GEX data
  ############
  DefaultAssay(seurat.human) <- "SCT"
  
  # create pseudobulk for visualisation
  pseudo.seurat <- AggregateExpression(seurat.human, 
                                   assays = "SCT",
                                   return.seurat = TRUE,
                                   group.by = c("seurat_clusters"))
  
  # set idents of pseudobulk object
  Idents(pseudo.seurat) <- "seurat_clusters"
  
  # extract top 10 DEGs per cluster
  goi <- seurat_clusters.gex.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::group_by(cluster) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    dplyr::pull(gene)
  
  # heatmap of top 10 degs per cluster
  DoHeatmap(pseudo.seurat, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + NoLegend()
  
  dev.copy(pdf, paste0(output.dir, "Heatmaptop_10_DEGs_seurat_clusters_pseudobulk.pdf"))
  dev.off()
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(seurat_clusters.gex.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- seurat_clusters.gex.de %>%
      dplyr::filter(p_val_adj < 0.05) %>%
      dplyr::filter(cluster == cluster.id[i]) %>%
      dplyr::top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.seurat, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE)
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk__seurat_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
}

```







